PLANETESIMAL ACCRETION AT SHORT ORBITAL PERIODS

### 2022-08-08 main.tex
More edits

In all cases, the simulations are evolved with a base timestep of 1.7
days, which corresponds to 3\% of an orbital dynamical time
$\sqrt{a^3/G M_{*}}$. Due to the vastly differing growth timescales in
each case, a simulation is stopped when the growth of the most massive
body flattens out. In figure \ref{fig:alpha_beta}, we show the a-e
distribution of bodies in the initial (blue) and final (orange)
snapshots from each of the 4 simulations. The size of the points
indicates the relative masses of the bodies. Only in the case of small
$\alpha$ does a residual population of dynamically hot planetesimals
develop. The lack of high eccentricity planetesimals in the large
$\alpha$ case suggests that most encounters result in accretion
instead of scattering. In the case of large $\beta$, the protoplanets
and remaining planetesimals end up in a dynamically cool state,
compared to the initial conditions.
%TRQ albeit for small alpha, only slightly cooler.
(Are the inelastic collisions playing a significant role here?)

@@

To investigate whether any of these planetesimal disks underwent
runaway growth, we examine the time evolution of the maximum and mean
masses in each case. The ratio $m_{max}/\left< m \right>$ is plotted
in figure \ref{fig:alpha_beta_evo}. On this plot, a positive slope
indicates that the growth rate is accelerating with mass, which is
evident at various times in all four cases. Even with a large value of
$\beta$, which means that the effective collision cross section is
very near the geometric value, runaway growth still appears to
operate. This seems to imply that, as bodies collide and grow, the
relative increase in their gravitational focusing factors, rather than
their absolute size, is what drives the system towards runaway
growth. Although larger encounter velocities will lengthen the growth
timescales, this mode of growth appears to be inevitable, so long as
gravity is the dominant force in the system.
% TRQ: also mention the negative slope?

Additionally, these results suggest that the value of $\alpha$, which is a function of only the initial conditions, controls the qualitative outcome of accretion. Across most of a planet-forming disk, $\alpha$ is small, and frequent gravitational encounters between the growing bodies will allow oligarchic growth to operate. In the dispersion-dominated regime, close encounters drive the stirring between planetesimals and embryos \citep{weidenschilling89, ida90}. When $\alpha$ is large, the Hill sphere of a body is no longer mostly empty space, and close encounters that would otherwise result in stirring instead result in accretion. In this regime, we observe that runaway growth still commences, but nearly all of the planetesimals are swept up by the forming protoplanets, rather than being scattered onto higher eccentricity orbits, where they would otherwise remain as a remnant of the early stages of planet formation \citep{kokubo98, kokubo00}.

@@

\begin{table}
% I'm going to replace the fdLo simulation with a lower resolution version m_pl = 5e22 for consistency
% Originally wanted to run everything at m_pl = 1e22, which corresponds to 100km bodies, but this wasn't feasible at higher Sigma
\caption{Summary of Full Disk Simulations Run}
\begin{tabular}{llllll} \hline \hline
Name    & $m_{pl}^{a}$ & $N_{pl}^{b}$ & $A^{c}$ & $\alpha^{d}$ &  \\ \hline
fdHi                 & $5 \times 10^{22}$ & 903,687 & 100                                                         & 1.5                              &   \\
fdHiShallow    & $5 \times 10^{22}$ & 903,687 & 100                                                        & 0.5                              &   \\
fdHiSteep       & $5 \times 10^{22}$ & 903,687 & 100                                                         & 2.5                              &   \\
fdLo                & $5 \times 10^{22}$ & 45,185    & 1                                                             & 1.5                              &   \\ \hline
\end{tabular}
\begin{flushleft}
\textsc{Note} \\ {$^a$ Planetesimal mass [g] \\
		       $^b$ Number of planetesimals \\
		       $^c$ Solid surface density normalization (relative to MMSN) \\
		       $^d$ Solid surface density power law index } \\
\end{flushleft}
\label{tab:sims}
\end{table}

@@

Motivated by the apparent dependence of accretion modes on $\alpha$,
we next investigate whether this highly efficient, non-oligarchic
growth mode should be expected to operate in a typical planet-forming
disk. Given the dearth (TRQ: why is this a reason for a wide disk?) of short period terrestrial planets observed around M stars (citation), we model a series of wide planetesimal disks, which span from 1 to 100 days in orbital period orbiting a late-type M star of mass 0.08 $M_{\odot}$. For a population of planetesimals with a bulk density of 3 g cm$^{-3}$, this orbital period range corresponds to $\alpha \in (0.7, 0.05)$. By simultaneously modeling a broad range of orbital periods, we can determine the critical value of $\alpha$ that divides these two modes of accretion, and also explore how the oligarchic/non-oligarchic accretion boundary affects the resulting distribution of protoplanets.

Four wide disk simulations are run in total (see table \ref{tab:sims}). In each case, the solid surface density follows a power law profile
\begin{equation}
	\Sigma(r) = \left( \frac{M_{*}}{M_{\odot}} \right) A \Sigma_{MMSN} \left( \frac{r}{1 AU} \right)^{-\alpha},
\end{equation}
%TRQ: should we use a different greek letter for the power law index?
where $M_{*}$ is the mass of the central star, $\Sigma_{MMSN} = 10$ g cm$^{-2}$ is surface density of the minimum-mass solar nebula \citep{hayashi81} at 1 AU, and $A$ is an enhancement factor. In the first case (fdHi), we model a disk that follows a MMSN power law slope, with the overall normalization enhanced by a factor of 100. This choice of parameters for the solid surface density profile appears necessary in order to reproduce many observed short period terrestrial worlds in-situ \citep{hansen12}. Additionally, we vary the power law index (fdHiShallow, fdHiSteep) and overall normalization (fdLo) of the solids profile.

>>>>

In all cases, the simulations are evolved with a base timestep of 1.7
days, which corresponds to 3\% of an orbital dynamical time
$\sqrt{a^3/G M_{*}}$. Due to the vastly differing growth timescales in
each case, a simulation is stopped when the growth of the most massive
body flattens out. In figure \ref{fig:alpha_beta}, we show the a-e
distribution of bodies in the initial (blue) and final (orange)
snapshots from each of the 4 simulations. The size of the points
indicates the relative masses of the bodies. Only in the case of small
$\alpha$ does a residual population of dynamically hot planetesimals
develop. The lack of high eccentricity planetesimals in the large
$\alpha$ case suggests that most encounters result in accretion
instead of scattering. In the case of large $\beta$, the protoplanets
and remaining planetesimals end up in a dynamically cool state,
compared with the initial conditions. This is due to kinetic energy 
being lost as particles inelastically collide. One final point to note 
is the difference between eccentricities of the protoplanets in the 
large $\alpha$, large $\beta$ and the small $\alpha$,
large $\beta$ case. The dynamically cooler result of the latter case
is likely due to the dominant role that inelastic collisions play here.

@@

To investigate whether any of these planetesimal rings underwent
runaway growth, we examine the time evolution of the maximum and mean
masses in each simulation. The ratio $m_{max}/\left< m \right>$ is plotted
in figure \ref{fig:alpha_beta_evo}. On this plot, a positive slope
indicates that the maximum and mean masses are diverging (i.e.
the growth rate is accelerating with mass), which is
evident for every case we have tested. Even with a large
$\beta$, which means that the effective collision cross section is
very near the geometric value, runaway growth still appears to
operate. The ubiquity of the early positive trends in this figure indicates
that, as bodies collide and grow the
relative difference in gravitational focusing factors between bodies
is what drives the system towards runaway
growth. Although larger encounter velocities lengthen the growth
timescales, this mode of growth appears to be inevitable, so long as
gravity is the dominant force in the system. For the large $\alpha$
cases, the curves in this figure eventually turn over and begin to decline.
Upon inspecting the simulation snapshots, we find that this trend is
driven by an increase in the average mass as the planetesimal population
becomes depleted. One should expect that the same thing will eventually
occur for small $\alpha$, although many more timesteps are required to
begin to run out of planetesimals.

Additionally, these results suggest that the value of $\alpha$, which is a function of only the initial conditions (the physical and Hill radius both scale with $M^{1/3}$, so $\alpha$ stays constant as bodies grow), controls the qualitative outcome of accretion. Across most of a planet-forming disk, $\alpha$ is small, and frequent gravitational encounters between the growing bodies will facilitate oligarchic growth. In the dispersion-dominated regime, close encounters drive the stirring between planetesimals and embryos \citep{weidenschilling89, ida90}. When $\alpha$ is large, the Hill sphere of a body is no longer mostly empty space, and close encounters that would otherwise result in stirring instead result in accretion. In this regime, we observe that runaway growth still commences, but nearly all of the planetesimals are swept up by the forming protoplanets, rather than being scattered onto higher eccentricity orbits, where they would otherwise remain as a remnant of the early stages of planet formation \citep{kokubo98, kokubo00}.

@@

\begin{table}
\caption{Summary of Full Disk Simulations Run}
\begin{tabular}{llllll} \hline \hline
Name    & $m_{pl}^{a}$ & $N_{pl}^{b}$ & $A^{c}$ & $\alpha^{d}$ &  \\ \hline
fdHi                 & $5 \times 10^{22}$ & 903,687 & 100                                                         & 1.5                              &   \\
fdHiShallow    & $5 \times 10^{22}$ & 903,687 & 100                                                        & 0.5                              &   \\
fdHiSteep       & $5 \times 10^{22}$ & 903,687 & 100                                                         & 2.5                              &   \\
fdLo                & $5 \times 10^{22}$ & 45,185    & 1                                                             & 1.5                              &   \\ \hline
\end{tabular}
\begin{flushleft}
\textsc{Note} \\ {$^a$ Planetesimal mass [g] \\
		       $^b$ Number of planetesimals \\
		       $^c$ Solid surface density normalization (relative to MMSN) \\
		       $^d$ Solid surface density power law index } \\
\end{flushleft}
\label{tab:sims}
\end{table}

@@

Motivated by the dependence of accretion modes on $\alpha$,
we next investigate whether this highly efficient, non-oligarchic
growth mode should be expected to operate in a typical planet-forming
disk. Given that N-body simulations of short-period terrestrial planet formation 
typically begin with a chain of neatly-spaced, isolation mass 
(see \citet{kokubo00} eq. 20) protoplanets, it is pertinent to determine 
whether the high $\alpha$ growth mode we revealed
in the previous section invalidates this choice of initial conditions.

 Given the dearth of short-period terrestrial planets observed around M stars (e.g. TRAPPIST-1), we chose to model a series of wide planetesimal disks, which span from 1 to 100 days in orbital period orbiting a late-type M star of mass 0.08 $M_{\odot}$. For a population of planetesimals with a bulk density of 3 g cm$^{-3}$, this orbital period range corresponds to $\alpha \in (0.7, 0.05)$. By simultaneously modeling a broad range of orbital periods, we can determine the critical value of $\alpha$ that divides these two modes of accretion, and also explore how the oligarchic/non-oligarchic accretion boundary affects the resulting distribution of protoplanets.

Four wide disk simulations are run in total (see table \ref{tab:sims}). In each case, the solid surface density follows a power law profile
\begin{equation}
	\Sigma(r) = \left( \frac{M_{*}}{M_{\odot}} \right) A \Sigma_{\mathrm{MMSN}} \left( \frac{r}{1 \mathrm{AU}} \right)^{-\delta},
\end{equation}
where $M_{*}$ is the mass of the central star, $\Sigma_{MMSN} = 10$ g cm$^{-2}$ is surface density of the minimum-mass solar nebula \citep{hayashi81} at 1 AU, and $A$ is an enhancement factor. In the first case (fdHi), we model a disk that follows a MMSN power law slope, with the overall normalization enhanced by a factor of 100. This choice of parameters for the solid surface density profile appears necessary in order to reproduce many observed short period terrestrial worlds in-situ \citep{hansen12}. Additionally, we vary the power law index (fdHiShallow, fdHiSteep) and overall normalization (fdLo) of the solids profile.
